Investigating Neolithic pottery usage from Guijiabao in the Yanyuan Basin in Southwest China using organic residue analysis

### 2023-02-04 analysis/paper/paper.Rmd
edited the background section

The recent growing application of organic residue analysis demonstrates its usefulness to understand ancient pottery use and function [@Kimpe2004; @Mayyas2022], human subsistence strategies or culinary practices across a region [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022]. Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers to trace their original foodstuffs [@Copley2005; @Whelton2021]. This is because lipid molecules can be accumulated and well preserved in the porous fabric of pottery over thousand years due to the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017]. In this study, we detect and identify lipid residues absorbed into the ceramic matrix that can reflect the original contents cooked or stored in a vessel during its use-life [@Dudd1999; @Miller2020].

@@

<!--Yanyuan area and Guijiabao site-->
We focus on Guijiabao site located in the Yanyuan Basin in Sichuan Province. Yanyuan Basin is a pull-apart basin located in the middle eastern part of the Hengduan Mountain, which is a group of parallel mountain ranges in a north-south direction from the Tibetan Plateau to the Yunnan-Guizhou Plateau (Figure \@ref(fig:SW-china-map)). River valleys of Hengduan Mountain provide pathways for the movements of people and the exchange of goods between cultural groups since the prehistoric period [@Lin2022; @Hao2022]. Among the archaeological sites in Yanyuan, Guijiabao (ca. 4800-4300 BP) located in the intersection of the middle Hengduan Mountain ranges, the Sichuan Basin, and the Yunnan-Guizhou Plateau is a unique site with continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b]. A survey in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site with long-term occupation [@Hao2022]. The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (ca. 5000-3700 BP), the Bronze Age (ca. 3200-2700 BP), and the late North Song to early Ming Dynasties (ca. 1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone [@Hao2022]. For the Neolithic period, it can be further separated into two subphases, 5000-4500 ca. BP and 4500-3700 ca. BP, according to the content of artifact assemblages and ceramic forms and styles.

<!--the major finding at GJB-->
At Guijiabao, ceramics with surface decorations such as incised geometric patterns and impressed dot patterns are the two common styles that have also been found in western Yunnan, indicative of frequent interactions between Guijiabao and sites in Yunnan [@Guijiabao2017a]. The archaeobotanical remains reveal crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on the analysis of flotation samples from ash pits and stratigraphic layers in 2015 [@Guijiabao2016]. The adoption of millet cultivation can be traced back to ca. 5000 BP [@Hao2022]. In addition, rice grains (*Oryza sativa*) have also been found in ash pits throughout different phases where the earliest rice dates to 4700 ca. BP [@Hao2022]. A detailed phytolith analysis for soil samples dating to Neolithic period (5000 ca. BP to 3700 ca. BP) further suggests the presence of rice phytoliths throughout the whole period in addition to millet phytoliths, indicative of a mixed millet/rice cultivation prior to 5000 ca. BP than previously thought [@Huan2022].

<!--info about Gaoshan-->
Gaoshan site is located in the southwestern margin of the Chengdu Plain, an alluvial plain between the Sichuan Basin and the Tibetan Plateau (Figure \@ref(fig:SW-china-map)). Excavations at Gaoshan revealed abundant pottery, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015]. According to the pottery forms and decorations, Gaoshan can be categorized into the earliest phase of the Baodun period, the earliest Neolithic culture (ca. 4700-3700 BP) in the Chengdu Plain [@Guedes2013]. It is proposed that Gaoshan could be even older than the Baodun culture according to a burial context, which was found in the stratigraphic layer earlier than the Baodun period [@Gaoshan2017]. A large quantity of carbonized rice grains was found at Gaoshan with a relatively small number of foxtail/broomcorn millets, suggesting rice is the main crop supplemented by millets [@JiangGaoshan2017]. This also indicates the spread of rice cultivation from the Middle and Lower Yangzi valley to Sichuan in southwest China occurred at least 4500 years ago [@Chi2010]. Moreover, recent research on animal bones in 2015 and 2016 shows that pigs and dogs were already domesticated, of which pig is the dominant animal [@He2020]. The primary subsistence strategy at Gaoshan could be rice cultivation with domesticated pigs, supplemented by millet cultivation.

 The current finding at Guijiabao and Gaoshan support this argument. However, plant remains from Guijiabao shows that millet is the major crop, which is at odds with the predominance of rice based on phytolich analysis and other macro-botanical analyses from Gaoshan or sites in Sichuan and Yunnan, such as Baodun, Henglanshan and Baiyangcun [@Hao2022; @Guedes2013]. Although the analysis of plant remains offers information on the adoption of crop cultivation and subsistence strategies, evidence directly reflects food preparation or processing using ceramics still lacks. This study aims to provide direct evidence from food residues in ceramics to investigate food choices associated with new practices of crop cultivation and local subsistence.

@@

## Samples
We selected five ceramic fragments from Guijiabao and one ceramic fragment from Gaoshan for organic residue analysis (Table XX). Four ceramic fragments from Guijiabao were recovered from trench 13 excavated in the 2017 field season and another sample was from the 2016 season. To understand the content and use of pottery vessels in Yanyuan in an earlier period, we analyzed potsherds from the stratigraphic layer which dates to the early Neolithic period (5000-4500 BP.) with potsherds from a later period (1000 BP.) for comparison. The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins. For the pottery types, cord-marked grey pottery vessels with fine inclusions and cord-marked sand-tempered vessels in reddish-brown color with coarse inclusions are the two dominant types [@Hao2022]. Our samples are jar fragments from those two common types that are supposed to be used for food serving or storage due to the archaeological context of settlement [@Hao2022; @Guijiabao2017a].

@@

The results show that most potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A). The potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well. A few number of potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003; @Dudd1998milk; @Suryanarayan2021]. Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020]. This also corresponds to the lipid distribution of the same potshard which reflects a possible plant origin based on the GC-MS results. Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for accessing food categories.

@@

# Discussion
<!--A brief summary of interesting results-->
Our results show that food choices indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were mainly dominated by ruminant fats during the Neolithic period. Moreover, it is striking to identify a non-ruminant source from a Guijiabao potshard that might originate from C3 plants. Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support that those ceramics were used as cooking or storage pots. The low yield might be related to the part of pottery samples analyzed in this study are mostly base fragments [@Evershed2008experimental; @Kwak2020]. But we also found a base sample dating to ca. 4500 BP. contains a rich composition of long-chain fatty acids, reflecting an excellent preservation. We recognize the limitation for interpreting food choices over time or differences across regions because of our small sample sizes. However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices or the local subsistence at Guijiabao.

>>>>

The recent growing application of organic residue analysis demonstrates its usefulness to understand ancient pottery use and function [@Kimpe2004; @Mayyas2022], human subsistence strategies, and culinary practices across a region [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022]. Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers to trace their original foodstuffs [@Copley2005; @Whelton2021]. This is because lipid molecules can be accumulated and well preserved in the porous fabric of pottery over thousand years due to the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017]. In this study, we detect and identify lipid residues absorbed into the ceramic matrix that can reflect the original contents cooked or stored in a vessel during its use-life [@Dudd1999; @Miller2020]. We attempt to explore the culinary practice under the circumstances during the time of frequent cultural interactions and expansion of agriculture by our case study.

@@

<!--Yanyuan area and Guijiabao site-->
We focus on Guijiabao (ca. 4800-4300 BP) located in the Yanyuan Basin in Sichuan Province (Figure \@ref(fig:SW-china-map)). Yanyuan Basin is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges. Guijiabao is especially unique due to its location as the intersection of the middle Hengduan Mountain ranges, the Sichuan Basin, and the Yunnan-Guizhou Plateau, with a continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b]. A survey in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site with long-term occupation [@Hao2022]. The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (ca. 5000-3700 BP), the Bronze Age (ca. 3200-2700 BP), and the late North Song to early Ming Dynasties (ca. 1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone [@Hao2022]. For the Neolithic period, it can be further separated into two subphases, 5000-4500 ca. BP and 4500-3700 ca. BP, according to the content of artifact assemblages and ceramic forms and styles.

<!--the major finding at GJB-->
The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins. Most ceramics have surface decorations, such as incised geometric patterns and impressed dot patterns that resemble ceramic decorations found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan [@Guijiabao2017a]. In addition, cord-marked pottery vessels are also commonly found in the early period. About the plant remains, the flotation analysis in 2015 reveal crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on samples collected from ash pits and stratigraphic layers [@Guijiabao2016]. The earliest domesticated millet grain can be traced back to ca. 5000 BP according to its radiocarbon date [@Hao2022]. A smaller number of rice grains (*Oryza sativa*) have also been found in ash pits where the earliest one dates to 4700 ca. BP. [@Hao2022]. Later, a detailed phytolith analysis for soil samples dating to the Neolithic period (5000 ca. BP to 3700 ca. BP) further suggests the presence of rice phytoliths throughout the temporal sequences, indicative of a mixed millet/rice cultivation prior to 5000 ca. BP. [@Huan2022].

<!--info about Gaoshan-->
To have a comparison, we also look at a sample from Gaoshan, which is located on the southwestern margin of the Chengdu Plain (Figure \@ref(fig:SW-china-map)). The Chengdu Plain is an alluvial plain between the Sichuan Basin and the Tibetan Plateau. Excavations at Gaoshan revealed abundant materials, including pottery, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015]. According to the pottery forms and decorations, Gaoshan can be categorized into the earliest phase of the Baodun period, the earliest Neolithic culture (ca. 4700-3700 BP) in the Chengdu Plain, or even earlier than that [@Guedes2013; @Gaoshan2017]. A large quantity of carbonized rice grains was found at Gaoshan with a relatively small number of foxtail/broomcorn millets, suggesting rice is the main crop supplemented by millets [@JiangGaoshan2017]. This also indicates the spread of rice cultivation from the Middle and Lower Yangzi valley to Sichuan in southwest China occurred at least 4500 years ago [@Chi2010]. Moreover, recent research on animal bones in 2015 and 2016 shows the predominance of domesticated pigs, with a small number of domesticated dogs [@He2020]. The primary subsistence strategy at Gaoshan could be rice cultivation with pig husbandry, supplemented by millet cultivation.

Although the current finding at Guijiabao and Gaoshan support the spread of mixed farming of millet/rice into southwest China around 5000 BP., what their proportions are in local culinary practices during the early time are unclear. For example, plant remains from Guijiabao show that millet is the major crop, which is at odds with the predominance of rice based on phytolith analysis. For Gaoshan and neighboring sites in Sichuan and Yunnan, such as Baodun, Henglanshan, and Baiyangcun, it appears that rice accounts for the vast majority of plaint remain [@Hao2022; @Guedes2013]. The analysis of plant remains offers information on the adoption of crop cultivation and subsistence strategies, but evidence directly reflects food preparation or processing using ceramics still lacks. This study aims to provide direct evidence from food residues in ceramics to investigate culinary practices associated with the cultivation of new crops.

@@

## Samples
We selected five ceramic fragments from Guijiabao and one ceramic fragment from Gaoshan for organic residue analysis (Table XX). Four ceramic fragments from Guijiabao were recovered from trench 13 excavated in the 2017 field season and one fragment was from the 2016 season. Those potsherds can be divided into two temporal phases, the early Neolithic period (5000-4500 BP.) and the historical period (1000 BP.). The common ceramic is cord-marked sand-tempered vessels unearthed at Guijiabao, and their forms include jars, urns, bowls, and basins [@Guijiabao2017a; @Hao2022]. Our samples are mostly reddish-brown jar fragments with medium-size sand temper and some of them has incised cord-marked decoration. Although the original function of those jars is uncertain, they should be supposed to be used for cooking or storage according to the char on the surfaces and their archaeological contexts from a settlement [@Hao2022; @Guijiabao2017a].

@@

The results show that most potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A). The potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well. A number of potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003; @Dudd1998milk; @Suryanarayan2021]. Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020]. This also corresponds to the lipid distribution of the same potshard which reflects a possible plant origin based on the GC-MS results. Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for accessing food categories.

@@

# Discussion
<!--A brief summary of interesting results-->
Our results show that food choices indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were mainly dominated by ruminant fats during the Neolithic period. Moreover, it is striking to identify a non-ruminant source from a Guijiabao potshard that might originate from C3 plants. Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support that those ceramics were used as cooking or storage pots. The low yield might be related to the part of pottery samples analyzed in this study are mostly base fragments [@Evershed2008experimental; @Kwak2020]. But we also found a base sample dating to ca. 4500 BP. contains a rich composition of long-chain fatty acids, reflecting a good preservation. We recognize the limitation for interpreting food choices over time or differences across regions because of our small sample sizes. Also, the lack of local modern references for animal fats and plants has impacts on our explanation. However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices or the local subsistence at Guijiabao.
